### 2016-11-12 Paper_v1.tex
Nov12

%The results are presented in \textbf{past~tense}. Each paragraph starts with a statement of the result in that paragraph in active voice.
%Each results paragraph ends with a Taken together type statement followed by a link statement e.g., Next we considered etc. When referring to figures, state what the figures shows (Fig. ZZ).

%\begin{enumerate}
%	\item{\textbf{First~section}:Description of the model biology}
%	\item{\textbf{Second~section}:Estimation of the model parameters, and refinement of the model structure (inclusion of the AA degradation pathways)}
%	\item{\textbf{Third~section}:Analysis of the flux distribution (over the ensemble?), sensitivity results (first parameters, then AA)}
%\end{enumerate}

%We generated mass balance equations around metabolites and enzymes in the network, modeled as ordinary differential equations with reaction rates equal to the product of a kinetic term and a control term. We used multiple saturation kinetics to model metabolic fluxes and mass action kinetics to model enzyme degradation.
%We modeled the control term using a rule-based approach in which each control factor had a regulatory transfer function, modeled as a Hill function, and the control term was calculated as the mean of all transfer functions.
%The model was trained against the training dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (ATP, ADP, AMP, GTP, GDP, GMP, CTP, CDP, CMP, UTP, UDP, UMP), and 18 of the 20 proteinogenic amino acids.
%The ensemble predicts most carbon flux going through glycolysis and the TCA cycle, and to a lesser extent the Entner-Doudoroff pathway  (Fig.~\ref{fig:Network}, bottom two rows of gray boxes). Pentose phosphate is much less utilized.
%We compared this flux profile with the steady-state fluxes calculated by sequence-specific FBA (Fig.~\ref{fig:Network}, top row of gray boxes).
%In the steady-state case, carbon flux avoids the upper reactions of glycolysis in favor of the Entner-Doudoroff pathway.
%It also not does participate in the TCA cycle, nor the reactions of pentose phosphate that are not required for Entner-Doudoroff.

\subsection*{Estimation of an ensemble of cell-free protein synthesis models.}
We used the hybrid cell-free modeling framework of Wayman et al. to simulate the production of a model protein [REFHERE].
The cell-free \textit{E. coli} metabolic model was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle),
as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids.
We generated an ensemble of N = 100 diverse parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model, and choosing sets so as to minimize the Pearson correlation coefficient between sets in the ensemble.

The ensemble of models captured the time evolution of cell-free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
The ensemble and the best-fit set also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:CarbonBoth}, top) and the totals by base, as well as the amino acid measurements (Fig.~\ref{fig:Amino}).
Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).

\subsection*{Sensitivity analysis}
We performed a local sensitivity analysis to determine the network reactions with the greatest effect on protein production and overall system state.
CAT production was msot sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis, as well as various reactions in glycolysis, the TCA cycle, and amino acid synthesis and degradation.

%We conducted a local sensitivity analysis to determine which of the kinetic and control parameters affected model performance.
%We calculated performance as area under the CAT curve, which was directly related to CAT synthesis rate, as the culture time was fixed and no CAT degradation was modeled.
%We randomly chose 180 sets of the 18,000 sets in the ensemble and defined these as our sub-ensemble; for each set in the sub-ensemble, we varied the rate constant and saturation constants of each metabolic flux and measured the resulting change in CAT production to estimate the sensitivity of model performance to that parameter.
%We did the same for the control parameters, both the order (Hill coefficient) and gain (related to the the dissociation constant).
%This allowed us to estimate the relative importance of the kinetic and control parameters to CAT production across the ensemble.
%Of the rate constants, those with the highest positive sensitivities were CAT synthesis, GTP synthesis, GMP synthesis, glutamine synthesis, and aspartate synthesis (Fig.~\ref{fig:LocalSensitivity}, top).
%This is explained by GTP and the amino acids being reactants for CAT synthesis.
%Also, GMP synthesis increases the total amount of guanosine, allowing for more GTP production.
%The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
%While GTP degradation is obvious, the others can be explained in that they consume ATP as well as several amino acids, all of which are reactants for CAT synthesis.
%Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
%This is because an increase in saturation constant causes a decrease in the corresponding reaction rate.
%The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
%Only two had a small standard error across the ensemble, relative to the ensemble mean sensitivity: the gain and order for pyruvate acting as an inhibitor on the pdh reaction.
%This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
%Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:CarbonBoth}), this suggests that allosteric control is not the limiting factor to CAT production.

%We conducted a global sensitivity analysis on the parameters that could be controlled experimentally: the initial conditions of glucose, oxygen, amino acids, and enzymes.
%We used the variance-based method of Sobol, and the same objective function of area under the CAT curve.
%Using a parameter set of relatively good fit against data, we defined parameter bounds and generated a Sobol sequence of 111,600 parameter values that fit within those bounds.
%We then calculated the total-order sensitivity and confidence interval for each of the experimentally controllable initial conditions.
%As the sensitivities were total-order, they were guaranteed to be non-negative.
%The largest sensitivities belonged to the initial conditions of the CAT macromolecular synthesis machinery, GTP synthase, GTP degradation, some amino acids such as phenylalanine, proline, and leucine, and some amino acid synthases (Fig.~\ref{fig:GlobalSensitivity}).
%This is all explained by GTP and amino acids being reactants for CAT synthesis.
%While some of the amino acids and amino acid synthases were among the highest in sensitivity, theirs were also very uncertain relative to the CAT macromolecular synthesis machinery and GTP synthase.
%The initial conditions of glucose and oxygen were among the least important according to the global sensitivity analysis, suggesting that the model predicts that CAT production can be sustained by consuming initial stores or can be powered by other means.

@@

\subsection*{Generation of model ensemble}
We generated an ensemble of 100 diverse parameter sets via a Markov chain Monte Carlo random walk.
Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-squared-error between experimental data and model predictions:
\begin{equation}\label{eqn:cost-function}
    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}_i}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
\end{equation}
where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes the weight of the $i$th dataset, $\mathcal{T}_i$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
We then perturbed model parameters:
\begin{equation}\label{eqn:parameter-perturbation}
    k_i^{new}=k_i*exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
\end{equation}
where $\mathcal{P}$ denotes the number of parameters, equal to 815, which includes 163 rate constants, 163 enzyme degradation rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, and $r_i$ denotes a random sample from the normal distribution.
For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
All sets with a lower cost than the previous set, and some with higher cost, were added to the ensemble.
After generating 12,437 sets, we selected 100 sets with minimal correlation to each other so as to avoid over-sampling any region of parameter space.
The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.93 between pairs of sets; the 100-set ensemble had a mean Pearson correlation coefficient of 0.84 between pairs of sets.

@@

Likewise, we determined the reactions most important to the system as a whole by computing the local sensitivity of the species for which data exist to each rate constant in the network.
Each sensitivity index was formulated as:
\begin{equation}\label{eqn:local-sensitivity}
\begin{split}
    S_{ij}=norm(X(p_i,p_j,t)-X(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
\end{split}
\end{equation}
where $S_{ij}$ denotes the sensitivity of the system state to the $i$th and $j$th parameters, and $X(p_i,p_j,t)$ denotes the system state, an array consisting of the concentration of every species for which data exist as a function of time and the $i$th and $j$th parameters.
% $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.

>>>>

\subsection*{Estimation of an ensemble of cell-free protein synthesis models.}
We used the hybrid cell-free modeling framework of Wayman et al. to simulate the production of a model protein [REFHERE].
The cell-free \textit{E. coli} metabolic model was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids.
We generated an ensemble of N = 100 diverse parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model, and choosing sets so as to minimize the Pearson correlation coefficient between sets in the ensemble.
We also defined the single set with lowest cost function as our best-fit set.

The ensemble of models captured the time evolution of cell-free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
The central carbon metabolism (Fig.~\ref{fig:CarbonBoth}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were all well captured by the ensemble and the best-fit set.
Allosteric control was seen to be particularly important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).

\subsection*{Sensitivity analysis}
We performed a local sensitivity analysis to determine the network reactions with the greatest effect on protein production and overall system state.
%CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis, as well as various reactions in glycolysis, pentose phosphate, the TCA cycle, and amino acid synthesis and degradation (Fig.~\ref{fig:Sensitivity}, top).
CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis (Fig.~\ref{fig:Sensitivity}, top, section A).
The 16 next most important reactions to CAT production (section B) came from various pathways across the network: 4 each from glycolysis, the TCA cycle, and amino acid synthesis/degradation; 2 from pentose phosphate; and 1 each from the Entner-Doudoroff pathway and the energy species reactions.
The pairwise sensitivities (off-diagonal elements) are often quite different from the corresponding first-order sensitivities (diagonal elements), and lead to some interesting outcomes.
For example, glutamine synthesis and arginine degradation are both among the most important reactions to CAT production (they rank 5th and 10th, respectively).
This is likely because they both affect the sensitive glutamine-glutamate balance; glutamine synthesis consumes glutamate, while arginine degradation produces it.
However, when both are perturbed, their combined effect on the model is very low, as the respective consumption and production of glutamate cancel out.

The system state as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase (pyruvate being consumed to produce lactate) (Fig.~\ref{fig:Sensitivity}, bottom, section F).
The 30 next most important reactions to the system state (section G) came from various pathways across the network: 8 from amino acid synthesis/degradation; 6 from glycolysis; 4 from TCA; 2 each from pentose phosphate, Entner-Doudoroff, energy species reactions, NAD species reactions, and small molecule transport; 1 from oxidative phosphorylation, and 1 pyrophosphatase.
The system state has even more pairwise sensitivities that differ from the corresponding first-order sensitivities and stand out as significant.
For example, the first-order effect of alanine synthesis is very high; it consumes both pruvate and glutamate, two key species in the network.
In addition, a handful of alanine synthesis pairwise sensitivities are high.
However, there are enough reactions that, when paired with alanine synthesis, have little effect on the model.
Malic enzyme is one of these, as it produces the pyruvate that alanine synthesis consumes.
Thus, the total-order alanine synthesis sensitivity is low, placing it at the very bottom of section I.
Another interesting result of this analysis is seen in the intersection of sections F and G with section J.
The 53 reactions in section J were turned off in the best-fit set (rate constants were set to 0); therefore, the perturbation of these reactions (multiplying the rate constant by 1.01) had no actual effect on the model.
Thus, all pairwise sensitivities with reactions in section J can essentially be considered first-order sensitivities for the other reactions.
Interestingly, the reactions in section F and several in section G showed most of thier highest sensitivities when paired with the "non-effects" of section J.
Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose uptake via GTP and CTP, fumarate reductase, and SO4 transport.
This suggests that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.

@@

\subsection*{Generation of model ensemble}
We generated an ensemble of 100 diverse parameter sets via a Markov chain Monte Carlo random walk.
Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-squared-error between experimental data and model predictions:
\begin{equation}\label{eqn:cost-function}
    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}_i}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
\end{equation}
where $\mathcal{D}$ denotes the number of datasets, equal to 37, $w_i$ denotes the weight of the $i$th dataset, $\mathcal{T}_i$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ denotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
We then perturbed model parameters:
\begin{equation}\label{eqn:parameter-perturbation}
    k_i^{new}=k_i*exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
\end{equation}
where $\mathcal{P}$ denotes the number of parameters, equal to 815, which includes 163 rate constants, 163 enzyme degradation rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, and $r_i$ denotes a random sample from the normal distribution.
For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
All sets with a lower cost than the previous set, and some with higher cost, were added to the ensemble.
After generating 12,437 sets, we selected 100 sets with minimal correlation to each other so as to avoid over-sampling any region of parameter space.
The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.

@@

Likewise, we determined the reactions most important to the system as a whole by computing the local sensitivity of the species for which data exist to each rate constant in the network.
Each sensitivity index was formulated as:
\begin{equation}\label{eqn:local-sensitivity}
\begin{split}
    S_{ij}=norm(X(p_i,p_j,t)-X(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
\end{split}
\end{equation}
where $S_{ij}$ denotes the sensitivity of the system state to the $i$th and $j$th parameters, and $X(p_i,p_j,t)$ denotes the system state, an array consisting of the concentration of every species for which data exist as a function of time and the $i$th and $j$th parameters.
% $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity}, bottom).
